Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-04-20 sources/bmc_template_submission/bmc_article.tex
MK edits page 3 and 4

Physcraper is implemented with Python and can also be run on the command line.
It currently consists of 12 modules.
%% , classes, methods and functions.%%
For testing and improving Physcraper's Python code syntax quality, we
used the Pylint software following instructions from their website \cite{pylintWeb}
 and manual \cite{pylintManual}, with a ``.pylintrc" configuration file.

@@

BLAST is performed using the `blastn` algorithm \cite{camacho2009blast}
implemented in BioPython's \cite{cock2009biopython} NCBIWWW module \cite{ncbiwww}
modified to accept an alternative BLAST address.
Each sequence in the alignment is BLASTed once against all DNA sequences in GenBank.
New sequences are excluded for analysis if they 1) are not in the search taxon;
2) have an e-value above the cutoff (default to 0.00001); 3) fall outside a min
and max length threshold, defined as the proportion of the average length without
gaps of all sequences in input alignment (default values of 80% and 120%,
respectively); 4) or if they are either identical to or shorter than an existing
sequence in the input alignment and they represent the same taxon in OpenTree or
NCBI taxonomy.
An arbitrary maximum number of randomly chosen sequences per taxon are allowed
(default to 5).

@@

A user is interested in phylogenetic relationships within the genus \textit{Ilex}. Commonly
known as "hollies", the genus encompasses between 400-700 living species, and is
the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering
plants.

An online literature review in June 2020 (Google scholar search for "ilex phylogeny")
reveals that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data publicly available \cite{gottlieb2005molecular, yao2020phylogeny}.
\cite{gottlieb2005molecular} made original tree and alignment available in TreeBASE
(study 1091 \cite{treebase1091}). This tree (Gottlieb2005 tree from now on) samples 41 species, is available from
OpenTree's Phylesystem (study pg\_2827 \cite{pg2827}),
and has been integrated into OpenTree's synthetic tree \cite{mrcaott68451ott89474}.
The most recent \textit{Ilex} tree \cite{yao2020phylogeny} (Yao2020 tree from now on) is available in
OpenTree's Phylesystem (study ot\_1984 \cite{ot1984}),
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the Yao2020 tree \cite{yao2020phylogeny} is the best sampled phylogeny available for
the genus \textit{Ilex}.
In order to showcase Physcraper's performance, we updated
the tree with the smallest number of tips (the Gottlieb2005 tree) with the expectation
to use the tree with the largest number of tips (the Yao2020 tree) as a standard
for comparison and verification of Physcraper's results.

>>>>

Physcraper is implemented with Python and can be run on a Python interactive session,
as a Python script or using the command line interface we developped for it.
It currently consists of 12 modules.
%% , classes, methods and functions.%%
For testing and improving Physcraper's Python code syntax quality, we
used the Pylint software following instructions from its website \cite{pylintWeb}
 and manual \cite{pylintManual}, with a ``.pylintrc" configuration file.

@@

BLAST is performed using the `blastn` algorithm \cite{camacho2009blast}
implemented in BioPython's \cite{cock2009biopython} NCBIWWW module \cite{ncbiwww}
modified to accept an alternative BLAST address.
Each sequence in the alignment is BLASTed once against all DNA sequences in GenBank.
New sequences are excluded for analysis if they 1) are not in the search taxon;
2) have an e-value above the cutoff (default to 0.00001); 3) fall outside a min
and max length threshold, defined as the proportion of the average length without
gaps of all sequences in input alignment (default values of 80\% and 120\%,
respectively); 4) or if they are either identical to or shorter than an existing
sequence in the input alignment and they represent the same taxon in OpenTree or
NCBI taxonomy.
An arbitrary maximum number of randomly chosen sequences per taxon are allowed
(default to 5).

@@

A user is interested in phylogenetic relationships within the genus \textit{Ilex}. Commonly
known as ``hollies", the genus encompasses between 400-700 living species, and is
the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering
plants.

An online literature review in June 2020 (Google scholar search for ``ilex phylogeny")
reveals that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data publicly available \cite{gottlieb2005molecular, yao2020phylogeny}.
\cite{gottlieb2005molecular} made original tree and alignment available in TreeBASE
(study 1091 \cite{treebase1091}). This tree (Gottlieb2005 tree from now on) samples 41 species, is available from
OpenTree's Phylesystem (study pg\_2827 \cite{pg2827}),
and has been integrated into OpenTree's synthetic tree \cite{mrcaott68451ott89474}.
The most recent \textit{Ilex} tree \cite{yao2020phylogeny} (Yao2020 tree from now on) is available in
OpenTree's Phylesystem (study ot\_1984 \cite{ot1984}),
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the Yao2020 tree \cite{yao2020phylogeny} is the best sampled phylogeny available for
the genus \textit{Ilex}.
In order to showcase Physcraper's performance, we updated
the tree with the smallest number of tips (the Gottlieb2005 tree) with the expectation
to use the tree with the largest number of tips (the Yao2020 tree) as a standard
for comparison and verification of Physcraper's results.
